---
title: "ChronoCause Pilot"
author: "Karl Mulligan, Erik Zeiner"
date: "2026-01-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# CSP preamble

# load packages
library(tidyverse)
library(ggbeeswarm)
library(brms)
library(tidyboot)
library(tidyjson)
library(tidybayes)
library(patchwork)
library(GGally)
library(cowplot)
library(BayesFactor)
library(aida)   # custom helpers: https://github.com/michael-franke/aida-package
library(faintr) # custom helpers: https://michael-franke.github.io/faintr/index.html
library(cspplot)
library(knitr)
library(ggplot2)
library(ggstance)

##################################################

# these options help Stan run faster
options(mc.cores = parallel::detectCores(),
        brms.backend = "cmdstanr")

# use the CSP-theme for plotting
theme_set(theme_csp())

# global color scheme from CSP
project_colors = cspplot::list_colors() |> pull(hex)

# setting theme colors globally
scale_colour_discrete <- function(...) {
  scale_colour_manual(..., values = project_colors)
}

scale_fill_discrete <- function(...) {
  scale_fill_manual(..., values = project_colors)
}

##################################################
```

# ChronoCause Pilot

This is an analysis pipeline for v.2 ChronoCause, using pilot data collected from Prolific on ($N = 152$).

```{r import, include=FALSE}

# set working directory if needed
# setwd("ADD_YOUR_PATH/CC-ChronoCause")
raw <- read.csv("01-experiments/01-exp-non-leakyUrns/pilot2/data/results01.csv")
glimpse(raw)
```

### Preprocessing

To address our main research questions, we need to recode the data in a few ways:

- In items with a delay, we need to code which ratings came from the earlier ball (`responseEarly`), and which came from the later ball (`responseLate`).
- We need general labels for specific code outcomes, like *solid-solid* (`combo`)
- We need long-form pivoted DataFrames in order to group by early/late (`full_long`) or left/right (`full_long_LR`).


```{r}
full <- raw

# define order for plotting purposes
full$delay <- factor(full$delay, levels = c("simult", "short", "medium", "long"))

# code early and late and the delta between the two ($\Delta$ of ‘causal rating of later-occurring event’ - ‘causal rating of earlier-occurring event’)
full <- full %>% mutate(responseEarly = ifelse(delayedUrn == "none", NA, ifelse(delayedUrn == "left", responseRight, responseLeft)))
full <- full %>% mutate(responseLate = ifelse(delayedUrn == "none", NA, ifelse(delayedUrn == "left", responseLeft, responseRight)))
full <- full %>% mutate(delta = responseLate - responseEarly)

# check comprehension and attention
full <- full %>% mutate(correctComp = ifelse(is.na(correctResponse), NA, correctResponse == response))
full <- full %>% mutate(correctAtt = ifelse(is.na(correctResponseAttention), NA, correctResponseAttention == responseAttention))


# filter out participants that somehow submitted more than once
# 32 is the number of rows per submission if participant goes through 2 rounds of comprehension checks
full_filtered <- full %>%
  group_by(prolific_pid) %>%
  filter(n() <= 32) %>%
  ungroup()


getComprehensionPass <- function(id) {

  participant <- full_filtered %>% filter(prolific_pid == id)

  comprehensionChecks <- participant %>%
    filter(trialType %in% c("comprehension-1", "comprehension-2")) %>%
    group_by(trialType) %>%
    summarise(all_correct = all(correctComp, na.rm = TRUE), .groups = "drop") %>%
    arrange(trialType) %>%
    pull(all_correct)

  trainingAttentionCheck <- participant %>%
    filter(trialType == "training-attention") %>%
    summarise(all_correct = sum(correctAtt), .groups = "drop") %>%
    pull(all_correct)

  criticalAttentionCheck <- participant %>%
    filter(trialType == "critical-attention") %>%
    summarise(all_correct = sum(correctAtt), .groups = "drop") %>%
    pull(all_correct)

  tibble(
    comprehensionCheck = if (comprehensionChecks[1]) TRUE else comprehensionChecks[2],
    trainingAttentionCheck = trainingAttentionCheck,
    criticalAttentionCheck = criticalAttentionCheck
  )
}
correct_df <- map_dfr(full_filtered$prolific_pid, getComprehensionPass)
full_filtered <- bind_cols(full_filtered, correct_df)


full_filtered %>%
  filter(!comprehensionCheck) %>%
  distinct(prolific_pid) %>%
  nrow()
full_filtered %>%
  filter(criticalAttentionCheck <3) %>%
  distinct(prolific_pid) %>%
  nrow()

full <- full_filtered %>%
  filter(criticalAttentionCheck >2)

full %>%
  distinct(prolific_pid) %>%
  nrow()
#
# df1 %>%
#   filter(!correctComp) %>%
#   distinct(prolific_pid) %>%
#   nrow()
#

# long-form dataframe for analyses grouping by early/late
full_long <- pivot_longer(full, cols = c("responseEarly", "responseLate"), names_to = "which_ball_EL", values_to = "causal_val")
# long-form dataframe for analyses grouping by left/right
full_long_LR <- pivot_longer(full, cols = c("responseLeft", "responseRight"), names_to = "which_ball_LR", values_to = "causal_val")

# conjunctive/disjunctive-only versions
conj <- full %>% filter(structure == "conjunctive")
disj <- full %>% filter(structure == "disjunctive")

conj_long <- full_long %>% filter(structure == "conjunctive")
disj_long <- full_long %>% filter(structure == "disjunctive")
conj_LR <- full_long_LR %>% filter(trialType == "critical" & structure == "conjunctive")
disj_LR <- full_long_LR %>% filter(trialType == "critical" & structure == "disjunctive")
```

## Main research question: does time (delay in events) affect how people attribute causality?

For each item, there is one of two outcomes: **win** or **lose**.

We solicit two ratings on a 7-point likert scale, from 1 (strongly disagree) to 7 (strongly agree):

- "Getting a {red/yellow/green/blue} ball caused Alice to {win/lose}."
- "Getting a {red/yellow/green/blue} ball caused Alice to {win/lose}."

We are interested in whether there is a difference in causal attribution between the earlier and the later events: the delta between these two ratings. We define $\Delta$: 'causal rating of the late-occurring event' -- 'causal rating of the earlier-occuring event'.

TODO: right now, the charts below are looking at averages across participants; this is wrong! We should be calculating deltas *per item*: make a new column for this.

### Conjunction hypotheses

**Comparison of mean responses across all participants**
```{r}
conj_long %>%
  filter(trialType == "critical") %>%
  filter(delay != "simult") %>%
  ggplot(aes(x = which_ball_EL, y = causal_val, color = delay)) +
  stat_summary(fun = 'mean', position = position_dodge()) +
  stat_summary(fun = 'mean', geom = 'line', aes(group = delay), position = position_dodge()) +
  stat_summary(fun.data = 'mean_se', position = position_dodge()) +
  geom_jitter(aes(), position = position_jitterdodge(jitter.width = 0.4,
                                                     jitter.height = .2,
                                                     dodge.width = 0.8),
              color = 'black', alpha = .1) +
  scale_y_continuous(breaks = 1:7) +
  facet_wrap(~combo) +
  scale_color_manual(values = c(project_colors[1], project_colors[3], project_colors[4]), name = 'delay length') +
  labs(title = "Conjunctive")
```

**Per-participant view**
```{r}

conj_long %>%
  filter(trialType == "critical") %>%
  filter(delay != "simult") %>%
  mutate(delay = factor(delay, levels = c("short", "long"))) %>%
  ggplot(aes(x = which_ball_EL, y = causal_val, color = delay, group = delay)) +
  geom_point(size = 3, position = position_dodgev(height = 0.2)) +
  geom_line(size = 1, position = position_dodgev(height = 0.2)) +
  scale_y_continuous(breaks = 1:7) +
  facet_grid(combo ~ as.integer(factor(prolific_pid))) +
  scale_color_manual(values = c(project_colors[1], project_colors[4]), name = 'delay length') +
  labs(title = "Conjunctive")
```

```{r}
conj_long %>%
  filter(trialType == "critical") %>%
  filter(delay != "simult") %>%
  ggplot(aes(x = delay, y = delta, fill = delay)) +
  stat_summary(fun = mean, geom = 'bar', width = 1.0) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.5, linewidth = 0.6, colour = project_colors[2]) +
  scale_fill_manual(values = c(project_colors[1], project_colors[3], project_colors[4]), name = 'delay length') +
  facet_wrap(~combo) +
  labs(y = 'Delta (late - early)', title = "Conjunctive")
```
#### Hypothesis 1: Categorical effect of time for conjunctive events
For conjunctive events, we expect there to be a preference for attributing causality to the *later* event in conditions with a (short or long) delay: i.e., a positive $\Delta$.

- 0 < $\Delta$(conjunctive, short) & 0 < $\Delta$(conjunctive, long)

The consequential conditions here are `solid-solid` (win) and `striped-striped` (lose).

SUPPORTED (valence effect for `striped-striped`?)

#### Hypothesis 2: Gradient effect of time for conjunctive events
For conjunctive events, we expect long delays to have a *greater* preference in causal attribution than short delays:

- 0 < $\Delta$(conjunctive, short) < $\Delta$(conjunctive, long)

The consequential conditions here are `solid-solid` (win) and `striped-striped` (lose).

UNSUPPORTED(?)

### Disjunction hypotheses

```{r}
disj_long %>%
  filter(trialType == "critical") %>%
  ggplot(aes(x = which_ball_EL, y = causal_val, color = delay)) +
  stat_summary(fun = 'mean', position = position_dodge()) +
  stat_summary(fun = 'mean', geom = 'line', aes(group = delay), position = position_dodge()) +
  stat_summary(fun.data = 'mean_se', position = position_dodge()) +
  geom_jitter(aes(), position = position_jitterdodge(jitter.width = 0.4,
                                                     jitter.height = .2,
                                                     dodge.width = 0.8),
              color = 'black', alpha = .1) +
  facet_wrap(~combo) +
  scale_y_continuous(breaks = 1:7) +
  scale_color_manual(values = c(project_colors[1], project_colors[3], project_colors[4]), name = 'delay length') +
  labs(title = "Disjunctive")
```

**Per-participant view**
```{r}
disj_long %>%
  filter(trialType == "critical") %>%
  filter(delay != "simult") %>%
  mutate(delay = factor(delay, levels = c("short", "long"))) %>%
  ggplot(aes(x = which_ball_EL, y = causal_val, color = delay, shape = delay, group = delay)) +
  geom_point(size = 3, position = position_dodgev(height = 0.2)) +
  geom_line(size = 1, position = position_dodgev(height = 0.2)) +
  scale_y_continuous(breaks = 1:7) +
  facet_grid(combo ~ as.integer(factor(prolific_pid))) +
  scale_color_manual(values = c(project_colors[1], project_colors[4]),
                     name = 'delay length') +
  labs(title = "Disjunctive")
```

```{r}
disj_long %>%
  filter(trialType == "critical") %>%
  filter(delay != "simult") %>%
  ggplot(aes(x = delay, y = delta, fill = delay)) +
  stat_summary(fun = mean, geom = 'bar', width = 1.0) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.5, linewidth = 0.6, colour = project_colors[2]) +
  scale_fill_manual(values = c(project_colors[1], project_colors[3], project_colors[4]), name = 'delay length') +
  facet_wrap(~combo) +
  labs(y = 'Delta (late - early)', title = "Disjunctive")
```
```{r}

disj_counts <- disj %>%
  filter(trialType == "critical") %>%
  filter(delay != "simult") %>%
  add_count(structure, combo, delay, responseEarly, responseLate, name = "n")
disj %>%

  ggplot(aes(x = responseEarly, y = responseLate))

```

#### Hypothesis 3: Categorical effect of time for disjunctive events
For conjunctive events, we expect there to be a preference for attributing causality to the *earlier* event in conditions with a (short or long) delay: i.e., a negative $\Delta$.

- 0 > $\Delta$(disjunctive, short) & 0 > $\Delta$(disjunctive, long)

The consequential conditions here are also `solid-solid` (win) and `striped-striped` (lose).

SUPPORTED

#### Hypothesis 4: Gradient effect of time for disjunctive events
For disjunctive events, we expect long delays to have a *greater* difference in causal attribution than short delays:

- 0 > $\Delta$(disjunctive, short) > $\Delta$(disjunctive, long)

The consequential conditions here are also `solid-solid` (win) and `striped-striped` (lose).

UNSUPPORTED

## Suggestions for next version

- Increase duration of `long` condition to potentially see a gradient effect
- Randomize order of rating prompts


